---
title: "Analyses for Nectar Larceny in the Trainbearers (Lesbia: Trochilidae)"
author: "Boris Igić"
date: "3/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> This document can be used to retrieve the basic analyses presented in the manuscript submission entitled "Nectar larceny in the trainbearers (Lesbia, Trochilidae)", by Boris Igic, Ivory Nguyen, and Phillip Fenberg.

```{r data-lib-import}
library(ggplot2)
library(RColorBrewer) # for brewer.pal() color palette
library(cowplot) # for plot_grid()

bandits<-read.csv(file = "/Users/boris/Dropbox/PROJECTS/nectar-robbing/larceny-paper/entered_data/robber_dbase_v1.1b-lesbia.csv", header=T, as.is=T)

# Splits away Lesbias
split.bandits <- split(bandits,bandits$bird.genus)
lesbias <- split.bandits$Lesbia
```

### Summary Statistics

Let's set up the catergories to be plotted. Two plots: one will include unknown modes and one will exclude them.

```{r plotting-setup-1}

# TOTAL RAW DATA WITH NA visits (Unknown mode)
lesbia.tally <- data.frame(modes=as.factor(c("R2","R1/R2","T","T/P","P","UNK")),
           obs=c(sum(lesbias$visitor.mode=="NR2",na.rm = T),
          sum(lesbias$visitor.mode=="NR",na.rm = T),
          sum(lesbias$visitor.mode=="T",na.rm = T),
          sum(lesbias$visitor.mode=="P/T",na.rm = T),
          sum(lesbias$visitor.mode=="P",na.rm = T),
          sum(is.na(lesbias$visitor.mode)))
          )
# This is the total tally for all Lesbias:
lesbia.tally

# Set factors into desired plotting order
lesbia.tally$modes <- factor(lesbia.tally$modes, levels = c("R2","R1/R2","T","T/P","P","UNK"))

# convert observations into percentages
lesbia.tally$perc <- (lesbia.tally$obs/sum(lesbia.tally$obs))*100
les.tally.perc<-lesbia.tally$perc
names(les.tally.perc)<-lesbia.tally$modes
round(les.tally.perc,2)

# TOTAL RAW DATA OMITTING NA visits (Unknown mode)
# This is intentionally clunky--helps separate what's going on
lesbia.noNA <- data.frame(modes=as.factor(c("R2","R1/R2","T","T/P","P")),
           obs=c(sum(lesbias$visitor.mode=="NR2",na.rm = T),
          sum(lesbias$visitor.mode=="NR",na.rm = T),
          sum(lesbias$visitor.mode=="T",na.rm = T),
          sum(lesbias$visitor.mode=="P/T",na.rm = T),
          sum(lesbias$visitor.mode=="P",na.rm = T)
          ))
# This is the total tally for all Lesbias:
lesbia.noNA

# Set factors into desired plotting order
lesbia.noNA$modes <- factor(lesbia.noNA$modes, levels = c("R2","R1/R2","T","T/P","P"))

# convert observations into percentages
lesbia.noNA$perc <- (lesbia.noNA$obs/sum(lesbia.noNA$obs))*100
les.noNA.perc<-lesbia.noNA$perc
names(les.noNA.perc)<-lesbia.noNA$modes
round(les.noNA.perc,2)

les.sp.not.na<-complete.cases(lesbias[,c(which(names(lesbias)=="bird.species"),which(names(lesbias)=="visitor.mode"))])
sp_mode <- lesbias[,c(1:16)][les.sp.not.na,]
nun <- split(sp_mode,sp_mode$bird.species)$nuna
vic <- split(sp_mode,sp_mode$bird.species)$victoriae
table(nun$visitor.mode)
table(vic$visitor.mode)
```

### Sex and Species

Now, let's check whether there are sex- or species-dependent associations. We don't expect any, a priori. To facilitate this, we're splitting the visits into two broad catergories: larceny (NR1,NR2,T, and any combinations thereof) and pollination (P,P/T). 

Why is P/T listed as a pollination? because we cannot rule it out as a pollination and that classification is conservative with respect to our main argument---there is a *lot* of larceny going on, about a half of all visits.

```{r lesbia-splits-v2,  eval=T}
lesbias$mode.bin<-lesbias$visitor.mode
lesbias$mode.bin[lesbias$visitor.mode=="NR"]<-"robbing"
lesbias$mode.bin[lesbias$visitor.mode=="NR2"]<-"robbing"
lesbias$mode.bin[lesbias$visitor.mode=="T"]<-"other"
lesbias$mode.bin[lesbias$visitor.mode=="T/P"]<-"other"
lesbias$mode.bin[lesbias$visitor.mode=="P/T"]<-"other"
lesbias$mode.bin[lesbias$visitor.mode=="P"]<-"other"

# Get rid of no species ID rows
split.lesbias <- split(lesbias,lesbias$bird.species)
victoriae <- split.lesbias$victoriae
nuna <- split.lesbias$nuna

lesbias.new <- rbind(nuna,victoriae)
lesbias.xtab <- xtabs( ~ bird.species + mode.bin + bird.sex, data = lesbias.new)

library(vcd)
library(vcdExtra)
library(MASS)

# simple summaries
# pdf(width=8,height=6,file="mosaic-plot.pdf")
mosaic(lesbias.xtab, shade=T)
# dev.off()
lesbias.table<-mosaic(lesbias.xtab, shade=T)

lesbias.table

assoc(lesbias.xtab, shade=TRUE)

# Statistical test of independence (variant repeated below for completeness)
summary(lesbias.xtab) # outputs chi-square test
```

## Model Testing


```{r lesbia-lm-stats, eval=T}

# Independent model
mod_ind<-loglm(~ mode.bin + bird.species + bird.sex, lesbias.xtab) 

# Partial Independence: mode is partially independent of composite variable sex*species
mod_part_ind1<-loglm(~ mode.bin + bird.species + bird.sex + bird.species * bird.sex, lesbias.xtab)

# Conditional Independence: mode is independent of species, given sex
mod_cond_1A<-loglm(~ mode.bin + bird.species + bird.sex + mode.bin * bird.sex + bird.species * bird.sex, lesbias.xtab)

# Conditional Independence: mode is independent of sex, given species
mod_cond_1B<-loglm(~ mode.bin + bird.species + bird.sex + mode.bin * bird.species + bird.sex * bird.species, lesbias.xtab)

# No Three-Way Interaction
mod_no3way_1<-loglm(~ mode.bin + bird.species + bird.sex + mode.bin * bird.species + bird.species * bird.sex + mode.bin * bird.sex, lesbias.xtab)

#anova(mod_ind,mod_part_ind1,mod_cond_1A,mod_cond_1B,mod_no3way_1)
#LRstats(mod_ind,mod_part_ind1,mod_cond_1A,mod_cond_1B,mod_no3way_1)
#anova(mod_ind,mod_part_ind1,mod_cond_1A,mod_no3way_1)
LRstats(mod_ind,mod_part_ind1,mod_cond_1A,mod_no3way_1)
```

### Set-up for plotting 

This sets up plots by species, but it is specifically omitting NA/UNK entries, because it is using the values of nun and vic as calculated for the Fisher Tests (which omit them).

```{r by-species-plotting-setup}
#
#L. nuna
#
nun.tally <- data.frame(modes=as.factor(c("R2","R1/R2","T","T/P","P","UNK")),
           obs=c(sum(nun$visitor.mode=="NR2",na.rm = T),
          sum(nun$visitor.mode=="NR",na.rm = T),
          sum(nun$visitor.mode=="T",na.rm = T),
          sum(nun$visitor.mode=="P/T",na.rm = T),
          sum(nun$visitor.mode=="P",na.rm = T),
          sum(is.na(nun$visitor.mode)))
          )
# This is the total tally for all Lesbias:
nun.tally


# Set factors into desired plotting order
nun.tally$modes <- factor(nun.tally$modes, levels = c("R2","R1/R2","T","T/P","P","UNK"))

# convert observations into percentages
nun.tally$perc <- (nun.tally$obs/sum(nun.tally$obs))*100

#
# L. victoriae
#

vic.tally <- data.frame(modes=as.factor(c("R2","R1/R2","T","T/P","P","UNK")),
           obs=c(sum(vic$visitor.mode=="NR2",na.rm = T),
          sum(vic$visitor.mode=="NR",na.rm = T),
          sum(vic$visitor.mode=="T",na.rm = T),
          sum(vic$visitor.mode=="P/T",na.rm = T),
          sum(vic$visitor.mode=="P",na.rm = T),
          sum(is.na(vic$visitor.mode)))
          )
# This is the total tally for all Lesbias:
vic.tally


# Set factors into desired plotting order
vic.tally$modes <- factor(vic.tally$modes, levels = c("R2","R1/R2","T","T/P","P","UNK"))

# convert observations into percentages
vic.tally$perc <- (vic.tally$obs/sum(vic.tally$obs))*100

```

### Plotting the joint species frequency of larceny

Two color schemes, red:blue:white and red:yellow:green, display all observations, including unknown visit modes (latter) and not including NA/UNK observations (former).

```{r lesbia-data-plot, fig.height = 10, fig.width = 5}

# Colors for plotting
# this is a good red, #f03b20, ("#FEEACCFF") but the result with white is very porky flesh. Looks like bacon
# '#e6550d' with '#FEEDDDFF' looks pretty good, because it's orangey
# red to blue? #f03b20 to #a6bddb
# scale_fill_manual(values=brewer.pal(n=6, name='RdYlGn')) <-- this is great, but
# it needs a sixth white/off-white color added for "Interaction cannot be determined"

##f03b20, ("#FEEACCFF"
lesbia.col <- colorRampPalette(c('#f03b20','cornflowerblue')) # reddish to blueish
lesbia.cols <- alpha(lesbia.col(5))
RBW6 <- c(lesbia.cols,'#EAEFEDFF')

# First plotting attempt with red:blue:white and observation percentages WITHOUT UNKNOWN mode observations
l1 <- ggplot(lesbia.noNA, aes(1, perc, fill=modes)) + 
  geom_bar(data=lesbia.noNA,stat="identity") +
  scale_fill_manual(values=RBW6) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), 
        axis.title.y=element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())

# Second plotting attempt with red:green:white, WITH UNKNOWN mode observations
#scale_fill_manual(values=brewer.pal(n=6, name='RdYlGn'))

RGW6 <- brewer.pal(n=6, name='RdYlGn')
RGW6[6] <- '#FAEFEFAF'
l2 <- ggplot(lesbia.tally, aes(1, perc, fill=modes)) + 
  geom_bar(data=lesbia.tally,stat="identity") +
  scale_fill_manual(values=RGW6) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), 
        axis.title.y=element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) 
  # + scale_y_reverse()

plot_grid(l1, l2, ncol=2, align="v")

# plot to file with pdf()
# pdf(file="lesbia-fig1.pdf", height=10, width=5)
# plot_grid(l1, l2, ncol=2, align="v")
# dev.off()
